Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-05-15 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex

\subsection{Order parameters from experiments}\label{expORDp}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{1}{2}\langle 3 \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^1$H-$^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions and also the connection between order parameters and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the effective dipolar coupling $d_\mathrm{CH}$ ($^1$H-$^{13}$C NMR) are calculated using equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7 kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
The effective dipolar coupling $d_\mathrm{CH}$ is related to the measured dipolar splitting $\Delta \nu_{CH}$ 
through scaling factor which depends on the pulse sequence used in the $^1$H-$^{13}$C NMR experiment~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion see Ref.~\citenum{accuracyPOST}.

The absolute values of order parameters are accessible with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. 
However, only $^1$H-$^{13}$C NMR techniques allow also the measurement of the sign of the order parameter~\cite{hong95a,hong95b,gross97}. 
The measured sign is negative for almost all the carbons discussed in this work, except for $\alpha$ which is positive~\cite{hong95a,hong95b,gross97}. 
For more detailed discussion about the determination of the sign of the order parameters see Ref.~\citenum{signpost}.

For most CH$_2$ segments in fluid phosholipid bilayer the order parameters are equal for both hydrogens attached to the same carbon.
However, in some cases (e.g. g$_1$, g$_3$ and  C$_2$ carbon in the \textit{sn}-2 chain) the order parameters are not equal and this 
can be observed with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. In the present work we call the phenomena of unequal order parameters 
for hydrogens attached to the same carbon as {\it forking} to avoid confusion with dipolar and quadrupolar splitting in NMR terminology. 
Forking has been studied in detail with $^2$H NMR techniques by deuterating the R or S position in CH$_2$ segment, and
the studies show that the forking arises from differently sampled orientations of the two C--H bonds, not from two 
separate populations of lipid conformations~\cite{engel81,gally81}.

@@

In general there is a good agreement between the order parameters measured with different experimental NMR techniques: Almost all the 
reported values are within a variation of $\pm$0.02 (which is also the error estimate given by Gross et al.~\cite{gross97}) 
for all fully hydrated PC bilayer, regardless of the variation in their acyl chain composition and temperature.
Exceptions are the somewhat lower order parameters sometimes reported from measurements using $^1$H-$^{13}$C NMR~\cite{hong95a,hong95b,warschawski05}.
These experiments are not shown in Fig.~\ref{HGorderparameters} as the reported error bars are either relatively large~\cite{hong95a,hong95b}, 
or the spectral resolution is quite low and the numerical lineshape simulations have not been used in the analysis~\cite{warschawski05}.
Due to this end, it is highly likely that these reported lower order parameters are due to lower experimental 
accuracy and therefore we exclude them from our discussion. 
For more details, see Ref.~\citenum{accuracyPOST}. 
Motivated by the high experimental reproducibility, we have highlighted in 
Fig.~\ref{HGorderparameters} the subjective sweet spots (light blue areas), within which we expect the calculated absolute 
values of the order parameters of a well-performing force field to fall.

@@

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^1$H-$^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured 
(see Ref.~\citenum{accuracyPOST} for more discussion.) 
However, as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. Consequently, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

\subsubsection{Phospholipid bilayer with low hydration level}
Fig.~\ref{ordPhydr} shows the published~\cite{dvinskikh05b,ulrich94,bechinger91} experimental order parameters
for the glycerol backbone and choline as a function of hydration level.
Despite slight differences in temperature and acyl chain composition,
the three independently reported data sets for the choline ($\beta$ and $\alpha$) segments agree well with each other:
Both order parameters increase with decreasing hydration level.
The glycerol backbone order parameters (g$_{3}$, g$_{2}$, g$_{1}$), in contrast,
have been observed~\cite{dvinskikh05b} to slightly decrease with dehydration.
Note that the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91} measured only 
absolute values, but here
we included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the negative $\beta$ order parameter actually increases with dehydration 
as its absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_lp.pdf}
%    \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_smooth.pdf}
    \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^1$H-$^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
    Note that to elucidate the relative change as a function of hydration level,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value at full hydration;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
  }
%  	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
%	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
%	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

@@

We note that the modest ($\Delta S_\mathrm{CH}<0.02$ for g$_1$; $<0.04$ for g$_2$, g$_3$; see Fig.~\ref{ordPchol})
effects of cholesterol
on the glycerol backbone order parameters of POPC
measured by $^1$H-$^{13}$C NMR~\cite{ferreira13}
agree well with the results for phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
This further supports the ideas that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup structure is largely independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_lp.pdf}
%    \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_smooth.pdf}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 model for g$_1$ are beoynd the y-axis scale.
    In order to elucidate the relative change as a function of cholesterol content,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value without cholesterol;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
}
%	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
%	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
%	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

@@

\subsection{Author Contributions}
{\it Alexandru Botan} provided simulation results for charmm36-UA\\
{\it Fernando Favela} prepared, performed and analyzed simulations
with popc and cholesterol for the CHARMM36 All-Atom FF. \\
{\it Patrick F. J. Fuchs} Ran and analyzed the Poger simulations. Provided scientific information which significantly advanced the project (signs and forking of order parameters). \\
{\it Matti Javanainen} prepared and performed simulations with multiple lipid models and analyzed the results. Supervised the work of JT.\\
{\it Matej Kanduc} provided simulation results for the Berger DLPC model \\
{\it Waldemar Kulig} prepared the MD simulations with cholesterol for the MacRog FF \\
{\it Antti Lamberg} Prepared and performed simulations with Berger and its variants to show the importance of the signs and stereospecific labeling of the order parameters. \\
{\it Claire Loison} provided simulation results for charmm36-UA. \\
{\it Alexander Lyubartsev} provided simulation results for Högberg08 force field.\\
{\it Markus S. Miettinen} Co-designed the project with OHSO and supported in the work management. Provided trajectories for Berger DMPC; prepared parameter files for 72 lipids \\
{\it Luca Monticelli} Critical discussions in all phases of the project. Collaboration with Jukka  M{\"a}{\"a}tt{\"a}. \\
{\it Jukka M{\"a}{\"a}tt{\"a}}  prepared and performed simulations with Berger and Slipids models and analyzed the results.\\
{\it O. H. Samuli Ollila}Co-designed the project with MSM and managed the work. Ran and analyzed several simulations. Wrote the manuscript. \\
{\it Marius Retegan} Prepared and validated the GROMACS compatible parameter files for GAFFlipid and Lipid14 force fields.\\
{\it Tomasz Rog} Provided the topologies for the MacRog model and also the full hydration simulation data with this model. \\
{\it Hubert Santuz} prepared and performed the cholesterol simulations with CHARMM36 and analyzed the results. 
Made significant contribution in the data management of the project.\\
{\it Joona Tynkkynen} prepared and performed the dehydration simulations with the MacRog FF.\\

>>>>

\subsection{Order parameters from experiments}\label{expORDp}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{1}{2}\langle 3 \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^1$H-$^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions, and also the connections between order parameter and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the effective dipolar coupling $d_\mathrm{CH}$ ($^1$H-$^{13}$C NMR) are calculated using the equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7~kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
The effective dipolar coupling $d_\mathrm{CH}$ is related to the measured dipolar splitting $\Delta \nu_\mathrm{CH}$ 
through a scaling factor that depends on the pulse sequence used in the $^1$H-$^{13}$C NMR experiment~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion, see Ref.~\citenum{accuracyPOST}.

The absolute values of order parameters are accessible with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. 
However, only $^1$H-$^{13}$C NMR techniques allow also the measurement of the sign of the order parameter~\cite{hong95a,hong95b,gross97}. 
The measured sign is negative for almost all the carbons discussed in this work, only the choline $\alpha$ is positive~\cite{hong95a,hong95b,gross97}. 
For more detailed discussion about the sign measurements of the order parameters, see Ref.~\citenum{signpost}.

For most CH$_2$ segments in a fluid phospholipid bilayer, the order parameters of both hydrogens are equal.
However, in some cases (e.g., g$_1$, g$_3$, and  C$_2$ carbon in the \textit{sn}-2 chain) the two order parameters are not equal;
this  can be observed with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. In the present work,
to avoid confusion with the dipolar and quadrupolar splittings in NMR terminology,
we call the phenomenon of unequal order parameters for hydrogens attached to the same carbon {\it forking}. 
Forking has been studied in detail with $^2$H NMR techniques by deuterating the R or S position in CH$_2$ segment, and
the studies show that it arises from differently sampled orientations of the two C--H bonds, not from two 
separate populations of lipid conformations~\cite{engel81,gally81}.

@@

In general there is a good agreement between the order parameters measured with different experimental NMR techniques: Almost all the 
reported values are within a variation of $\pm$0.02 (which is also the error estimate given by Gross et al.~\cite{gross97}) 
for all fully hydrated PC bilayer, regardless of the variation in their acyl chain composition and temperature.
Exceptions are the somewhat lower order parameters sometimes reported from measurements using $^1$H-$^{13}$C NMR~\cite{hong95a,hong95b,warschawski05}.
These experiments are not shown in Fig.~\ref{HGorderparameters} as the reported error bars are either relatively large~\cite{hong95a,hong95b}, 
or the spectral resolution is quite low and the numerical lineshape simulations have not been used in the analysis~\cite{warschawski05}.
Due to this end, it is highly likely that these reported lower order parameters are due to lower experimental 
accuracy and therefore we exclude them from our discussion. 
For more details, see Ref.~\citenum{accuracyPOST}. 
Motivated by the high experimental repeatability, we have highlighted in 
Fig.~\ref{HGorderparameters} the subjective sweet spots (light blue areas), within which we expect the calculated absolute 
values of the order parameters of a well-performing force field to fall.

@@

\subsection{Response to dehydration and cholesterol content}
In addition to pure phosphatidylcholine bilayers at full hydration, the choline headgroup order parameters
have been measured under various different conditions~\cite{gally75,brown77,brown78,akutsu81,altenbach84,scherer89,bechinger91,ulrich94,dvinskikh05b,castro08,kapla12,ferreira13}.
Also the order parameters for the glycerol backbone have been measured with $^1$H-$^{13}$C NMR in dehydrated conditions~\cite{dvinskikh05b}, and as a function 
of anesthetics~\cite{castro08} and glycolipids~\cite{kapla12} for DMPC, and as a function of cholesterol 
concentration for POPC~\cite{ferreira13}. Due to the high resolution in the NMR (especially $^2$H NMR) experiments,
even very small order parameter changes resulting from the varying conditions can be measured 
(see Ref.~\citenum{accuracyPOST} for more discussion.) 
However, as already discussed above, it is not simple to deduce 
the structural changes from order parameter changes~\cite{akutsu91,semchyschyn04}. Consequently, comparison of the order parameters
between simulations and experiments in different conditions can be used to assess the quality of the force field 
in different situations, and, if the quality is good, to potentially interpret the structural changes in experiments.
Here we exemplify such comparison for a lipid bilayer under low hydration levels and when varying amounts of cholesterol is included in the bilayer. 
The interaction between ions and a phosphatidylcholine bilayer will be discussed in a separate study~\cite{ionpaper}.

\subsubsection{Phospholipid bilayer with low hydration level}
Fig.~\ref{ordPhydr} shows the published~\cite{dvinskikh05b,ulrich94,bechinger91} experimental order parameters
for the glycerol backbone and choline as a function of hydration level.
Despite slight differences in temperature and acyl chain composition,
the three independently reported data sets for the choline ($\beta$ and $\alpha$) segments agree well with each other:
Both order parameters increase with decreasing hydration level.
The glycerol backbone order parameters (g$_{3}$, g$_{2}$, g$_{1}$), in contrast,
have been observed~\cite{dvinskikh05b} to slightly decrease with dehydration.
Note that the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91} measured only 
absolute values, but here
we included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the negative $\beta$ order parameter actually increases with dehydration 
as its absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_lp.pdf}
    \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_smooth.pdf}
    \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^1$H-$^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
    Note that to elucidate the relative change as a function of hydration level,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value at full hydration;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
  }
  	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

@@

We note that the modest ($\Delta S_\mathrm{CH}<0.02$ for g$_1$; $<0.04$ for g$_2$, g$_3$; see Fig.~\ref{ordPchol})
effects of cholesterol
on the glycerol backbone order parameters of POPC
measured by $^1$H-$^{13}$C NMR~\cite{ferreira13}
agree well with the results for phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
This further supports the ideas that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup structure is largely independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_lp.pdf}
    \includegraphics[width=7.3cm]{../DATAreportediINblog/cholesterolization_smooth.pdf}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 model for g$_1$ are beoynd the y-axis scale.
    In order to elucidate the relative change as a function of cholesterol content,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value without cholesterol;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
}
	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

@@

\subsection{Author Contributions}
{\it Alexandru Botan} provided simulation results for charmm36-UA\\
{\it Fernando Favela} prepared, performed and analyzed simulations
with popc and cholesterol for the CHARMM36 All-Atom FF. \\
{\it Patrick F. J. Fuchs} Ran and analyzed the Poger simulations. Provided scientific information which significantly advanced the project (signs and forking of order parameters). \\
{\it Matti Javanainen} prepared and performed simulations with multiple lipid models and analyzed the results. Supervised the work of JT.\\
{\it Matej Kanduc} provided simulation results for the Berger DLPC model \\
{\it Waldemar Kulig} prepared the MD simulations with cholesterol for the MacRog FF \\
{\it Antti Lamberg} Prepared and performed simulations with Berger and its variants to show the importance of the signs and stereospecific labeling of the order parameters. \\
{\it Claire Loison} provided simulation results for charmm36-UA. \\
{\it Alexander Lyubartsev} provided simulation results for Högberg08 force field.\\
{\it Markus S. Miettinen} \todo{Please write a text summarizing your contribution} \\
{\it Luca Monticelli} Critical discussions in all phases of the project. Collaboration with Jukka  M{\"a}{\"a}tt{\"a}. \\
{\it Jukka M{\"a}{\"a}tt{\"a}}  prepared and performed simulations with Berger and Slipids models and analyzed the results.\\
{\it O. H. Samuli Ollila} Designed and managed the work. Ran and analyzed several simulations. Wrote the manuscript.  \\
{\it Marius Retegan} Prepared and validated the GROMACS compatible parameter files for GAFFlipid and Lipid14 force fields.\\
{\it Tomasz Rog} Provided the topologies for the MacRog model and also the full hydration simulation data with this model. \\
{\it Hubert Santuz} prepared and performed the cholesterol simulations with CHARMM36 and analyzed the results. 
Made significant contribution in the data management of the project.\\
{\it Joona Tynkkynen} prepared and performed the dehydration simulations with the MacRog FF.\\
